---
title: "Data processing"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_depth: 5
    number_sections: true
---

# Data Processing
In this section, we transform data by adding some new calculated columns, clean dirty data and prepare it for analysis.

## Utility functions
First we define a function used along the file to display sample of data calculated and another one to return the memory size of data frame.

```{r sample display, warning=FALSE, results='hide', message=FALSE}
# A function to display a sample of data
sample <- function(df, columns, n) {
  selected_df <- df %>%
    select(all_of(columns))
  
  unique_combinations <- df %>%
    distinct(across(all_of(columns)), .keep_all = TRUE)
  
  sampled_df <- unique_combinations %>%
    slice_sample(n = n)
  select(sampled_df, columns)
}

# A function to display memory size of data frame
size <- function(df) {
  size_in_bytes <- object.size(df)
  size_in_mb <- size_in_bytes / 1048576
  return(size_in_mb)
}
```

## Times columns calculation
We need to calculate some new columns about different time units

```{r Modify local settings to english, echo=FALSE, warning=FALSE, results='hide', message=FALSE}
# Set the United States as the regional parameter to use English names for months and days.
Sys.setlocale("LC_TIME", "en_US.UTF-8")
```

### Calculate month
Month of trips is calculated from start date.

```{r Month calculation, warning=FALSE, message=FALSE}
df$month <- month(df$started_at, label = TRUE)

sample(df, all_of(c("started_at", "month")), 10)
```

### Calculate day of week
Week day of trips is calculated from start date. We also define an order on values of day column.

```{r day calculation, warning=FALSE, message=FALSE}
df$day <- weekdays(df$started_at)

# define an order on days
df$day <- factor(df$day, levels = c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday", "Sunday"))

sample(df, all_of(c("started_at", "day")), 10)
```

### Calculate hour of trip start
Hour column is calculated from start date.

```{r Hour calculation, warning=FALSE, message=FALSE}
df$hour <- hour(df$started_at)

sample(df, all_of(c("started_at", "hour")), 10)
```

### Calculate date of trip start
Date of trip is calculated from start date, the time part is ignored. This column is used to make daily analysis.

```{r date calcualtion, warning=FALSE, message=FALSE}
df$date <- date(df$started_at)

sample(df, all_of(c("started_at", "date")), 10)
```

### Calculate season of trip
Season of trip is calculated from month: Spring: "Mar", "Apr", "May"; Summer: "Jun", "Jul", "Aug"; Autumn: "Sep", "Oct", "Nov"; Winter: "Dec","Jan", "Feb";

```{r season calculation, warning=FALSE, message=FALSE}
df <- df %>%
  mutate(
    season = case_when(
      month %in% c("Mar", "Apr", "May") ~ "spring",
      month %in% c("Jun", "Jul", "Aug") ~ "summer",
      month %in% c("Sep", "Oct", "Nov") ~ "autumn",
      TRUE ~ "winter" # Décembre, janvier, février
    )
  )

sample(df, all_of(c("month", "season")), 10)
```

### Calculate duration column
We calculate duration of each trip.

```{r duration column, warning=FALSE, results='hide', message=FALSE}
df$duration <- difftime(df$ended_at, df$started_at, units = "mins")

sample(df, all_of(c("started_at", "ended_at", "duration")), 10)
```

## Geographic columns calculation
Here we calculate two new columns about distance of trips.

### Calculate distance of trips
Straight-line distance is calculated from start and end points coordinates, using Haversine formula.

```{r distance column, warning=FALSE, message=FALSE}
df$distance <- ifelse(
  !is.na(df$start_lat) & !is.na(df$start_lng) & !is.na(df$end_lat) & !is.na(df$end_lng),
  distHaversine(
    cbind(df$start_lat, df$start_lng),
    cbind(df$end_lat, df$end_lng)
  )/1000,NA
)
sample(df, all_of(c("start_lat", "start_lng", "end_lat", "end_lng", "distance")), 10)
```

### Calculate distance category
We calculate distance categories labeled: [0, 1km], ]1km, 3km], ]3km, 10km], ]10km, 50km] and "Not defined". Last category contains distance greater than 50km.

```{r distance calculation, warning=FALSE, message=FALSE}
# Categorize the trips based on distance
df <- df %>%
  mutate(dist_cat = case_when(
    distance <= 1 ~ "0<d<1km",
    distance > 1 & distance <= 3 ~ "1km<d<3km",
    distance > 3 & distance <= 10 ~ "3km<d<10km",
    distance > 10 & distance <= 50 ~ "10km<d<50km",
    TRUE ~ "Abnormal distance" # For distances that don't fall into the specified ranges
  ))
# define an order on dist_cat values.
df$dist_cat<- factor(df$dist_cat, levels = c("0<d<1km", "1km<d<3km", "3km<d<10km", "10km<d<50km", "Abnormal distance"))

sample(df, all_of(c("distance", "dist_cat")), 10)
```

## Reorder columns and suppression
Here we delete unused columns and reorder new calculated columns.

### Delete rows with negative duration or distance \> 50km
**Important observation:**  
Transformations deleting or modifying data must be validated by the manager. 

Impossible trips with negative duration ('ended_at < started_at') holds for 272 rows. There may have been a technical error stemming from the synchronization of the internal timekeeping mechanism.

```{r negative duration supression}
original_nrow <- nrow(df)
df <- subset(df, duration >= 0)
ignored_rows <- original_nrow - nrow(df)

cat("Number of rows ignored, trips with negative duration:", ignored_rows, "\n")
```

Only three rows hold, we don't delete rows missing distance because they are due to missing end coordinates.

```{r distance > 50km supression}
original_nrow <- nrow(df)
df <- subset(df, distance <= 50 | is.na(distance))
ignored_rows <- original_nrow - nrow(df)

cat("Number of rows ignored, trips with too high distance:", ignored_rows, "\n")
```

### Reorder columns and delete unused ones
Modify the order of columns.

```{r Reorder columns}
df <- df %>%
  relocate(season, date, month, day, hour, duration, distance, dist_cat, .after = ended_at)

sample(df, all_of(c("season", "date", "month", "day", "hour", "duration", "distance", "dist_cat")), 10)
```

Delete unused columns to decrease size of manipulated data set. Following columns are deleted: *started_at, ended_at, start_station_name, start_station_id, end_station_name, end_station_id, end_lat, end_lng*
```{r Delete unused columns}
original_size <- size(df)

df <- df %>% 
      select(-c(ride_id, started_at, ended_at, end_station_id, end_station_name, start_station_id))

new_size <- size(df)

cat("Original size before suppression was ", original_size, " Mo, and become ", new_size, " Mo", " after suppression: ")
```


```{r Save processed in CSV, echo=FALSE, warning=FALSE, results='hide', message=FALSE}
# write_csv(df, "df023_processed.csv", quote = "all")
```
